---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # Chunks are shown rather than run: building this README requires a compiled
  # Rust toolchain, and a README that cannot be regenerated everywhere is worse
  # than one without printed output.
  eval = FALSE
)
```

# icebergr

When an organisation's data moves into Apache Iceberg, R stops being a
first-class citizen and becomes the thing you export CSVs to.

**R is the only major data language without an Apache Iceberg client.** Apache
governs implementations in Java, Python (PyIceberg), Rust and Go. There is none
for R. Until now the only route was to read Iceberg tables through DuckDB as an
intermediary, which means no writes, no schema access, no snapshot management,
no catalog integration and no partition information.

Meanwhile Snowflake, Databricks, BigQuery, AWS and Dremio have all standardised
on Iceberg as the open table format. Parquet is well served in R by `arrow`, and
Delta Lake has a community Rust binding; Iceberg had nothing.

`icebergr` talks to Iceberg directly, through
[`iceberg-rust`](https://github.com/apache/iceberg-rust), the Apache-governed
Rust implementation, via `extendr`. Arrow is the interchange layer throughout, so
data crosses from Rust into R over the Arrow C stream interface without a
serialisation round trip.

## Installation

CRAN is the intended destination; `icebergr` is not there yet. See
[CRAN status](#cran-status) for what has to clear first. Until then, install from
GitHub.

```{r}
# install.packages("pak")
pak::pak("PursuitOfDataScience/icebergr")
```

Installing from source compiles Apache Iceberg's Rust implementation, so you need
a Rust toolchain (`rustc` >= 1.92):

```sh
curl --proto '=https' --tlsv1.2 -sSf https://sh.rustup.rs | sh
```

Expect the first build to take a while: a default install compiles 264 crates.

Optional backends are off by default because each substantially enlarges that
tree:

```sh
# Object storage (S3)
ICEBERGR_CARGO_FEATURES=s3 R CMD INSTALL --preclean .

# AWS Glue catalog (implies s3)
ICEBERGR_CARGO_FEATURES=glue R CMD INSTALL --preclean .
```

Either of those needs network access, even from the CRAN tarball. The vendored
crates bundled there cover the default build; the roughly one hundred extra
crates these features pull in — the AWS SDK among them — would have tripled the
tarball, so cargo fetches them from crates.io instead.

## Getting started

Everything below runs offline against a locally generated table.

```{r}
library(icebergr)

tbl <- icebergr_example_table()
tbl
#> <icebergr_table>
#>   table:    db.events
#>   location: /tmp/RtmpXXXXXX/icebergr-warehouse/db/events
#>   format:   v2
#>   snapshot: 4964299904926817223
#>   columns:  5
#>     id <int>
#>     event <string>
#>     amount <double>
#>     day <date>
#>     recorded_at <timestamptz>
```

Read it, pushing the filter and the projection down into the scan:

```{r}
icebergr_collect(
  icebergr_scan(tbl, filter = id > 1000 & amount > 900, select = c("id", "amount"))
)
```

Pushdown is the whole performance argument for Iceberg over reading raw Parquet,
so it is worth being able to *verify* rather than assume. `icebergr_scan_plan()`
shows which files a scan would touch, before reading any of them:

```{r}
nrow(icebergr_scan_plan(icebergr_scan(tbl)))
#> [1] 2
nrow(icebergr_scan_plan(icebergr_scan(tbl, filter = id > 1000)))
#> [1] 1
```

Travel back through snapshot history:

```{r}
history <- icebergr_snapshots(tbl)
history[, c("snapshot_id", "operation", "added_records")]

# The state before the most recent append
icebergr_collect(icebergr_scan(tbl, snapshot_id = history$snapshot_id[[1]]))

# Or by time. `as_of` is a *commit* time, not a time in the data, so take it
# from the history: this table was written just now.
icebergr_collect(icebergr_scan(tbl, as_of = history$timestamp[[1]]))
```

Append new data:

```{r}
tbl <- icebergr_append(tbl, data.frame(
  id = 9001L,
  event = "purchase",
  amount = 42.5,
  day = as.Date("2024-07-01"),
  recorded_at = as.POSIXct("2024-07-01 09:00:00", tz = "UTC")
))
```

Connect to a real catalog. Credentials come from the environment, never from
arguments:

```{r}
Sys.setenv(ICEBERGR_REST_TOKEN = "...")
catalog <- icebergr_catalog("rest", uri = "https://catalog.example.com")

icebergr_list_namespaces(catalog)
icebergr_list_tables(catalog, "analytics")

tbl <- icebergr_table(catalog, "analytics.events")
icebergr_schema(tbl)
icebergr_partitions(tbl)
```

See `vignette("catalog-configuration")` for REST, Glue and S3 setup.

## Supported Iceberg features

Built against **`iceberg-rust` 0.10.0**. Reads and writes table spec **v1 and
v2**. v3 metadata is parsed, but v3-specific features are not exposed.

`icebergr_spec_support()` reports this matrix for your own build, resolved
against the optional features you compiled in.

✅ works &nbsp;·&nbsp; ⚙️ needs a build flag &nbsp;·&nbsp; 🦀 missing upstream in
`iceberg-rust` &nbsp;·&nbsp; 🚧 out of scope for 0.1.0

**📖 Reading**

| | |
| :-: | --- |
| ✅ | Read a table into Arrow / a tibble |
| ✅ | **Predicate pushdown** — filters prune whole files and row groups |
| ✅ | **Projection pushdown** — only selected columns leave disk |
| ✅ | Row group and row-level scan pruning |
| ✅ | Merge-on-read tables — positional *and* equality deletes applied. `iceberg-rust` refuses an equality delete keyed on a `list` or `map` column; every other column type is handled |
| ✅ | Inspect the file plan before reading, with `icebergr_scan_plan()` |
| ✅ | Nested types — `struct` and `list` read and write. A `map` column can be created, but writing map *values* from R needs the `arrow` package, since `nanoarrow` cannot build a map array alone. Note `nanoarrow::na_map()` needs its key type built non-nullable: `na_map(na_string(nullable = FALSE), …)` |
| 🦀 | Row `limit` pushdown — `limit` bounds decoding, not planning |
| 🦀 | Pushdown *on* a nested field — read the parent column and filter in R |
| ✅ | `decimal` filters — row-level selection is disabled for these scans, because `iceberg-rust` 0.10.0 drops every row of an ordering comparison on a decimal; file and row group pruning still apply |

**🕰️ Time travel**

| | |
| :-: | --- |
| ✅ | Snapshot history |
| ✅ | Read a snapshot by id |
| ✅ | Read as of a timestamp — resolved against Iceberg's snapshot log, so a rollback's abandoned snapshot is not selected |
| ✅ | Read the schema as of a snapshot, and filter and select by the names *it* had |

**✍️ Writing**

| | |
| :-: | --- |
| ✅ | Append rows |
| ✅ | Create a table (unpartitioned) and a namespace |
| ✅ | Register an existing table — from a metadata file named `<version>-<uuid>.metadata.json`, as every engine writes them; a renamed file reads but cannot be appended to |
| ✅ | Spec **v1** tables read, filter and append, and stay v1 |
| 🚧 | Appending to a *partitioned* table — refused before anything is written, since a partition value per row is not computed. Reading one works |
| 🦀 | Row-level deletes and `MERGE` / upsert — *reading* such tables works. `iceberg-rust` can write an equality delete file but has no transaction action that commits one |
| 🦀 | Overwrite writes — `fast_append` is the only way `iceberg-rust` 0.10.0 can add files |
| 🚧 | Partitioned table creation — available upstream, not exposed here |

**🗂️ Catalogs and metadata**

| | |
| :-: | --- |
| ✅ | REST catalog |
| ✅ | In-process `memory` catalog, for local warehouses |
| ✅ | Schema and partition spec inspection, including a partitioned table's spec |
| ✅ | Table properties, and `icebergr_reload()` to see another session's commits |
| 🚧 | Setting table properties — needs an `update_properties` transaction |
| ⚙️ | AWS Glue — build with the `glue` Cargo feature |
| ⚙️ | Object storage (S3) — build with the `s3` Cargo feature |
| 🦀 | Hadoop / filesystem catalog — none exists upstream; use `type = "memory"` |

**🚧 Not in 0.1.0** — schema evolution, partition evolution, snapshot expiry,
`dbplyr` lazy verbs, table encryption. Compaction is 🦀: it needs a rewrite action
`iceberg-rust` 0.10.0 does not have.

A correct narrow surface beats a broad buggy one. Anything marked 🦀 or 🚧 raises
an informative error rather than failing obscurely.

## Type fidelity

R types survive the Arrow round trip as follows:

| R type | Iceberg type | Round trip |
| --- | --- | --- |
| `integer` | `int` | Unchanged |
| `double` | `double` | Unchanged, including `Inf` and `-Inf`. **`NaN` returns as `NA`**: R's `is.na(NaN)` is `TRUE`, so it is written as a null. |
| `character` | `string` | Unchanged, UTF-8 preserved |
| `logical` | `boolean` | Unchanged |
| `Date` | `date` | Unchanged |
| `POSIXct` | `timestamptz` | Instant preserved; normalised to UTC |
| `bit64::integer64` | `long` | Unchanged, full 64-bit precision. Reading a `long` back needs `bit64` installed; without it Arrow's `int64` narrows to a `double`. |
| `factor` | `string` | **Returns `character`.** Iceberg has no dictionary type, so levels cannot be carried. |
| `POSIXct` | `timestamp_ns` | Readable, writable and filterable, but a `POSIXct` is a double of *seconds*, so sub-microsecond precision is lost. `nanoarrow` warns on every such read — it triggers on the nanosecond count exceeding 2^53, which any present-day instant does, so the warning appears even when nothing was actually lost. |
| data frame column | `struct` | Unchanged, and comes back as a data frame column. Iceberg cannot push a filter or a projection down *onto* a nested field, so read the parent column and subset it in R. |
| `vctrs::list_of` | `list` | Unchanged, and comes back as a `list_of` column. |

Snapshot ids are **character**, not numeric. Iceberg assigns them as random
64-bit integers and an R numeric holds only 53 bits, so a double round trip would
silently select the wrong snapshot.

## CRAN status

**CRAN is the target.** The package is built to submit: `configure` and
`configure.win`, `Makevars.in` templates, vendored dependencies via
`tools/vendor.R`, an offline `--offline` build, `-j2` to stay inside CRAN's
parallelism limit, a `CARGO_HOME` confined to the build tree, and a per-crate
`LICENSE.note` inventory. CI runs that exact path on every commit, so the
submission build is exercised continuously rather than assembled at the last
minute.

Two things have to clear first. Both are tractable; neither is settled.

**1. Vendored size — measured, not estimated.** The source tarball is **10.7 MB**,
almost all of it `vendor.tar.xz`. That is over CRAN's 10 MB guideline, which
explicitly allows requesting more; the request is in
[`cran-comments.md`](https://github.com/PursuitOfDataScience/icebergr/blob/main/cran-comments.md),
made with the arithmetic shown rather than asserted.

It was 31.5 MB until `tools/vendor.R` learned to reduce the vendor tree rather
than just compress it. `cargo vendor` writes every entry in `Cargo.lock`, and a
lockfile is a union over every platform and every optional feature, so most of
what it writes is never compiled: of 442 crates, **270** are compiled on at least
one platform R runs on and **264** on any single machine. The other 172 — the
optional Glue and S3 backends, and the crates belonging to operating systems R
does not run on — cannot be *deleted*, because cargo resolves the whole lock graph
before it filters targets or selects features and fails outright on a missing
package. But it never reads the ones it does not select, so they ship as their
manifest and their licence rather than as source. With that plus four smaller
sweeps (test corpora, files cargo ignores, `#[cfg(test)]` modules inside `src/`,
and the 219 of 246 `windows-sys` API modules behind disabled features), 373.8 MB
of vendored sources become 97.3 MB.

What is left is not reducible without giving something up. `iceberg` declares
`[features] default = []` and has no optional dependencies, so `tokio`, `reqwest`,
`parquet`, twelve `arrow-*` crates and `apache-avro` are unconditional. Compressed
individually the ten largest crates are 4.45 MB of the 10.46 MB archive and the
remaining 432 directories average 22 KB, so there is no longer a single item to
attack.

One consequence for anyone enabling an optional backend: `ICEBERGR_CARGO_FEATURES`
builds fetch their extra crates from crates.io, because the bundled archive covers
the default build. A default install still touches the network at no point, and CI
runs that offline path on every commit.

**2. Minimum Rust version.** `iceberg-rust` 0.10.0 declares `rustc` 1.94 and
edition 2024, and bumps its minimum most releases. This one is now measured
rather than assumed: CRAN's Windows build farm carries 1.92.0, and the first
submission failed to install there because of it.

The declared floor is not a real one. Four crates in the tree ask for 1.94 —
`iceberg`, `iceberg-catalog-rest`, `iceberg-catalog-glue` and `fastnum`, with
nothing else above 1.91.1 — and none of them uses a language or library feature
newer than 1.92. The whole tree compiles, and the test suite passes, on 1.92.0.
Cargo treats a dependency's `rust-version` as a hard error rather than a warning,
so `src/Makevars{,.win}` pass `--ignore-rust-version` and `tools/msrv.R` gates on
the floor the package is genuinely tested against, which `DESCRIPTION` states as
1.92.

Pinning an earlier `iceberg-rust` stays the fallback for the day a release really
does need something newer:

| `iceberg-rust` | Declared MSRV | Cost of pinning it |
| --- | --- | --- |
| 0.10.0 (current) | 1.94, builds on 1.92 | — |
| 0.9.1 | 1.92 | Loses `with_runtime`; the runtime is then inherited from the calling context, which is where we already are |
| 0.8.0 | 1.88 | Predates the storage-factory refactor; needs real binding changes |

Edition 2024 itself only needs 1.85, which is early 2025, so the edition is not
the constraint — the rolling minimum is.

**Precedent worth knowing.** `polars` was on CRAN in 2023 with a non-vendored,
network-fetching build that current policy disallows, and is on r-universe today.
That is a caution about *how* a heavy Rust package fails a submission, not proof
that one must.

[`FEASIBILITY.md`](https://github.com/PursuitOfDataScience/icebergr/blob/main/FEASIBILITY.md) has the full analysis, the measurements behind these numbers, and
what remains unverified.

## Licence

GPL (>= 3). Bundled Rust crates keep their own licences, listed in `inst/NOTICE` and
`LICENSE.note`.

Apache, Apache Iceberg and Iceberg are trademarks of The Apache Software
Foundation. `icebergr` is an independent community package, not affiliated with
or endorsed by the ASF, and is not one of the official Iceberg clients.
