---
title: "Catalog configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Catalog configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # Connection recipes cannot be executed here: they need catalog servers and
  # credentials, and a vignette that reaches the network is neither reproducible
  # nor permitted on CRAN.
  eval = FALSE
)
```

Connection recipes for the catalogs `icebergr` supports, how credentials are
handled, and what to do when a connection does not work. Code in this vignette is
shown but not run.

```{r}
library(icebergr)
```

## Credentials are never arguments

`icebergr` reads credentials from environment variables and provides no argument
to pass one. This is deliberate. A token passed as an argument ends up in the
script that called it, in `.Rhistory`, in any knitr cache of the chunk that ran
it, and in the `traceback()` of any error raised nearby.

| Variable | Iceberg property | Used for |
| --- | --- | --- |
| `ICEBERGR_REST_TOKEN` | `token` | Bearer token for a REST catalog |
| `ICEBERGR_REST_CREDENTIAL` | `credential` | OAuth2 client credential |
| `ICEBERGR_REST_OAUTH2_SERVER_URI` | `oauth2-server-uri` | OAuth2 token endpoint |
| `ICEBERGR_REST_SCOPE` | `scope` | OAuth2 scope |
| `ICEBERGR_S3_ACCESS_KEY_ID` | `s3.access-key-id` | Object storage access key |
| `ICEBERGR_S3_SECRET_ACCESS_KEY` | `s3.secret-access-key` | Object storage secret |
| `ICEBERGR_S3_SESSION_TOKEN` | `s3.session-token` | Temporary session token |
| `ICEBERGR_S3_REGION` | `s3.region` | Object storage region |
| `ICEBERGR_S3_ENDPOINT` | `s3.endpoint` | Custom or S3-compatible endpoint |

The standard `AWS_ACCESS_KEY_ID`, `AWS_SECRET_ACCESS_KEY`, `AWS_SESSION_TOKEN`
and `AWS_REGION` variables are consulted as a fallback, so an environment already
configured for AWS works without further setup.

Set them outside your scripts — in `~/.Renviron`, in your shell profile, or from
your platform's secret manager:

```
# ~/.Renviron
ICEBERGR_REST_TOKEN=eyJhbGciOi...
ICEBERGR_S3_REGION=us-east-1
```

Two guarantees worth relying on:

- `print()` on a catalog never shows its properties.
- Errors raised while connecting list property *keys* only, never values.

You can still pass a credential through `...` if you must, but `icebergr` warns
when you do, because it is nearly always a mistake.

## Iceberg REST catalog

The common case: Polaris, Lakekeeper, Nessie, Unity Catalog's Iceberg endpoint,
Tabular-style services, or a self-hosted REST catalog.

```{r}
Sys.setenv(ICEBERGR_REST_TOKEN = "...") # better: set this in ~/.Renviron

catalog <- icebergr_catalog(
  "rest",
  uri = "https://catalog.example.com/api/catalog",
  warehouse = "analytics"
)

icebergr_list_namespaces(catalog)
icebergr_list_tables(catalog, "analytics")
```

`warehouse` is whatever the server expects to identify the warehouse — often a
name rather than a path.

### With OAuth2 client credentials

```{r}
Sys.setenv(
  ICEBERGR_REST_CREDENTIAL = "client_id:client_secret",
  ICEBERGR_REST_OAUTH2_SERVER_URI = "https://auth.example.com/oauth/token",
  ICEBERGR_REST_SCOPE = "catalog"
)

catalog <- icebergr_catalog("rest", uri = "https://catalog.example.com")
```

### Extra, non-secret properties

Anything `iceberg-rust` accepts can be passed through `...`:

```{r}
catalog <- icebergr_catalog(
  "rest",
  uri = "https://catalog.example.com",
  `rest.signing-region` = "us-east-1",
  `rest.sigv4-enabled` = "true"
)
```

## Object storage

REST catalogs usually serve tables that live in S3. That needs the optional `s3`
Cargo feature, which is off by default because opendal is a substantial subtree:

```sh
ICEBERGR_CARGO_FEATURES=s3 R CMD INSTALL --preclean .
```

Check before relying on it:

```{r}
"s3" %in% icebergr_spec_support()$cargo_features
```

```{r}
Sys.setenv(
  ICEBERGR_S3_ACCESS_KEY_ID = "...",
  ICEBERGR_S3_SECRET_ACCESS_KEY = "...",
  ICEBERGR_S3_REGION = "us-east-1"
)

catalog <- icebergr_catalog(
  "rest",
  uri = "https://catalog.example.com",
  storage = "s3"
)
```

### S3-compatible storage (MinIO, R2, Ceph)

```{r}
Sys.setenv(
  ICEBERGR_S3_ENDPOINT = "https://minio.internal:9000",
  ICEBERGR_S3_ACCESS_KEY_ID = "...",
  ICEBERGR_S3_SECRET_ACCESS_KEY = "..."
)

catalog <- icebergr_catalog(
  "rest",
  uri = "https://catalog.internal/api/catalog",
  storage = "s3",
  `s3.path-style-access` = "true"
)
```

Most S3-compatible services need path-style access.

## AWS Glue

Needs the `glue` Cargo feature, which implies `s3`:

```sh
ICEBERGR_CARGO_FEATURES=glue R CMD INSTALL --preclean .
```

```{r}
catalog <- icebergr_catalog(
  "glue",
  warehouse = "s3://my-bucket/warehouse",
  `glue.region` = "us-east-1"
)

icebergr_list_namespaces(catalog)
```

Glue uses the AWS SDK's own credential chain, so instance roles, IRSA, SSO
profiles and `AWS_PROFILE` all work as they do elsewhere.

If the feature is not compiled in, the error says so and how to get it, rather
than failing at link time.

## Local warehouses

For local files, testing, or a warehouse on a mounted filesystem, use the
in-process `memory` catalog. This needs no server and no credentials, and is what
this package's own tests run against.

```{r}
warehouse <- "/data/warehouse"
catalog <- icebergr_catalog("memory", warehouse = warehouse)

icebergr_create_namespace(catalog, "db")
tbl <- icebergr_create_table(catalog, "db.events", data.frame(id = integer()))
```

There is deliberately no `"hadoop"` catalog type: **`iceberg-rust` does not
implement a Hadoop or filesystem catalog.** `memory` is the local equivalent.

Its one real limitation is that the table registry lives in memory, so it does
not persist between sessions. Tables already on disk are re-attached with
`icebergr_register_table()`:

```{r}
catalog <- icebergr_catalog("memory", warehouse = "/data/warehouse")
icebergr_create_namespace(catalog, "db")

tbl <- icebergr_register_table(
  catalog, "db.events",
  "/data/warehouse/db/events/metadata/00003-....metadata.json"
)
```

Point it at the *newest* metadata file: Iceberg writes a new one per commit, and
the newest is the current state of the table.

## Performance notes

- **Push filters down.** `filter` and `select` are the difference between reading
  a table and reading the part of it you want. Confirm with
  `icebergr_scan_plan()`.
- **`limit` is not pushdown.** `iceberg-rust` has no row limit in its scan API, so
  `limit` bounds decoding, not planning. To read less, filter.
- **Batch size trades memory for call overhead.** `batch_size` controls rows per
  Arrow batch; the default is usually right.
- **Worker threads.** The tokio runtime uses 2 worker threads, which keeps
  parallelism within CRAN's limits for checks. For large scans over object
  storage, raise it before loading the package:

  ```r
  Sys.setenv(ICEBERGR_WORKER_THREADS = "8")
  library(icebergr)
  ```

  The runtime starts on first use, so this has no effect once you have connected.
- **Reads stream.** Batches are pulled one at a time rather than collected up
  front, so memory stays bounded by batch size, not table size — until
  `icebergr_collect()` materialises the result into a tibble, which is necessarily
  all of it. To process a table larger than memory, scan it in filtered pieces.

## Troubleshooting

**`could not connect to the REST catalog`, with a list of property keys.** The
keys tell you what the catalog was given. A missing `token` usually means the
environment variable is not visible to R — check `Sys.getenv("ICEBERGR_REST_TOKEN")`,
and remember that `~/.Renviron` is read only at startup.

**TLS failures behind a corporate proxy.** `icebergr` uses rustls with the
platform trust store, so a CA installed in the system store is honoured. A CA
present only in R's own bundle is not.

**`is not available in this build of icebergr`.** An optional Cargo feature is
missing. The message names the feature and gives the install command.

**`The compiled Rust component of icebergr is not available.`** The package
loaded but its native routines did not. This is a half-finished source install;
reinstall and read the build log for cargo errors.

**Reads succeed, writes fail with a permissions error.** Iceberg commits write
both data files and new metadata. Write access to the table's `metadata/`
prefix is required, not just to `data/`.

**A table shows fewer rows than expected.** Reading a merge-on-read table is
supported: `iceberg-rust` applies positional and equality deletes during the
scan, so rows another engine deleted will correctly be absent. If the count still
looks wrong, check whether you are reading an older snapshot — `print()` on a
table shows which one — and whether a filter is pruning more than you meant, with
`icebergr_scan_plan()`.

**Writes to a table another engine performs deletes on.** `icebergr` only
appends, which is always safe. It cannot delete or update rows, so a workflow
needing that must do it elsewhere; see `icebergr_spec_support()`.

**`cannot append to "db.events": it is partitioned by …`.** Reading a partitioned
table works, and `icebergr_partitions()` reports its spec, but appending to one
would have to compute a partition value for every row, which this version does
not do. The append is refused before any data file is written, so nothing is left
behind in the warehouse. Write to such a table with an engine that supports
partitioned writes.
