---
title: "Predicate and Projection Pushdown"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
bibliography: vignette_reference.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Predicate and Projection Pushdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(icebergr)
```

Pushdown is the reason to read an Iceberg table through a client rather than
collect it and subset in R. `filter` and `select` travel with the scan into
`iceberg-rust`, which uses them to decide what to open — so the saving is in
bytes never read, not in rows discarded afterwards.

This vignette shows the three levels at which that happens, and how to verify
each of them rather than assume it.

```{r}
tbl <- icebergr_example_table(rows = 200)
```

## The three levels

| Level | Uses | Effect |
| --- | --- | --- |
| File | per-column bounds in the Iceberg manifest | a file is never opened |
| Row group | statistics in the Parquet footer | a block within a file is skipped |
| Row | the filter, re-applied during decode | non-matching rows are dropped |

The first two are *pruning* and cost nothing: the bounds were written at commit
time. Only the third touches the data. The equivalent measurements for column
stores in general are in @abadi2008column, and for Parquet and ORC specifically,
with their present-day encodings, in @zeng2023columnar.

## File pruning, measured

`icebergr_scan_plan()` returns the file tasks a scan would execute, without
executing them. Comparing two plans is how you confirm a predicate was actually
pushed down:

```{r}
all_files <- icebergr_scan_plan(icebergr_scan(tbl))
hot_files <- icebergr_scan_plan(icebergr_scan(tbl, filter = id > 1000))

nrow(all_files)
nrow(hot_files)
```

The example table is built with two appends, so there are two data files and the
second holds the higher ids. A filter that only the second can satisfy plans one
file instead of two:

```{r}
all_files[, c("record_count", "file_size_in_bytes")]
hot_files[, c("record_count", "file_size_in_bytes")]
```

`record_count` is the count *before* filtering — it is what the manifest recorded
when the file was written. A plan is a statement about what will be opened, not
about what will come back.

A predicate no file can satisfy prunes everything:

```{r}
nrow(icebergr_scan_plan(icebergr_scan(tbl, filter = id > 10^6)))
```

## Projection pushdown

`select` never reaches the row level at all, because a columnar file lets the
reader seek to the columns it wants and ignore the rest. This is Dremel's layout
[@melnik2010dremel], and it is why the column list matters as much as the
predicate:

```{r}
narrow <- icebergr_collect(
  icebergr_scan(tbl, select = c("id", "amount"), limit = 3)
)
narrow
```

The other columns are not decoded, not converted, and not allocated in R.

## Which predicates push down

`filter` is an R expression, translated into an Iceberg predicate. What is
translatable is a closed list — `==`, `!=`, `<`, `<=`, `>`, `>=`, `&`, `|`, `!`,
`%in%`, `is.na()`, `is.nan()` and `startsWith()`:

```{r}
icebergr_collect(
  icebergr_scan(tbl,
    filter = event %in% c("purchase", "refund") & amount > 900,
    select = c("event", "amount"), limit = 5
  )
)
```

A bare name is read as a column when the table has one of that name, and
otherwise evaluated in the calling environment — so a local variable works
without any quoting ceremony:

```{r}
threshold <- 950
icebergr_collect(
  icebergr_scan(tbl, filter = amount > threshold, select = "amount", limit = 3)
)
```

Anything outside that list belongs in R, after collecting. There is no partial
credit: an expression the translator does not recognise is an error rather than a
silently unpushed filter, because a filter that quietly stopped pruning would look
like nothing more than a slow scan.

## Three things that do not push down

**`limit` bounds decoding, not planning.** `iceberg-rust` 0.10.0 has no row limit
in its scan API, so every file the predicate admits is still planned:

```{r}
nrow(icebergr_scan_plan(icebergr_scan(tbl, limit = 1)))
```

The rows are counted as batches arrive, which does save conversion into R, but not
I/O. Use a predicate to reduce I/O.

**A filter on a nested field.** Iceberg cannot push a predicate onto a field
inside a `struct`; read the parent column and subset it in R.

**A `decimal` ordering comparison, at the row level only.** It is pushed down, but
with `iceberg-rust`'s row-level selection turned off for that scan, because in
0.10.0 that stage drops every row of an ordering comparison against a decimal.
File and row-group pruning still apply, so the scan is slightly less selective and
still correct. `icebergr_spec_support()` records this as a known upstream
limitation rather than leaving it to be discovered:

```{r}
features <- icebergr_spec_support()$features
features[
  grepl("decimal|limit", features$feature, ignore.case = TRUE),
  c("feature", "supported", "reason")
]
```

## Case sensitivity

Iceberg column names are case-sensitive, and `case_sensitive = FALSE` prefers an
exact match rather than taking the first hit. A table holding both `id` and `ID`
resolves each to itself; a name that matches two columns and neither exactly is an
error rather than a silent choice.

```{r}
icebergr_collect(icebergr_scan(tbl,
  select = "ID", case_sensitive = FALSE,
  limit = 2
))
```

## Reading a plan for cost, not just count

`file_size_in_bytes` is what the manifest recorded, so a plan gives an I/O
estimate before any read:

```{r}
sum(all_files$file_size_in_bytes)
sum(hot_files$file_size_in_bytes)
```

On object storage that ratio is the one that matters, and it is available without
a single `GET` against the data. Query engines built around this format use the
same metadata for the same purpose [@behm2022photon].

# References
