---
title: "What Iceberg Is, and What It Fixes"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
bibliography: vignette_reference.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What Iceberg Is, and What It Fixes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(icebergr)
```

A table format is not a file format. Parquet says how one file is laid out;
Iceberg says which files constitute a table right now, what their schema is, and
what the table looked like an hour ago. This vignette is about that distinction,
because it is the reason reading Parquet is not the same as reading an Iceberg
table — and the reason `arrow` cannot substitute for a client.

## The problem inherited from Hive

The original convention, from Hive [@thusoo2009hive], was that a table *is* a
directory, and a partition *is* a subdirectory. Everything follows from that: to
know what a table contains, list the filesystem.

Three consequences made it unworkable at scale, and they are what the lakehouse
table formats were designed against [@armbrust2021lakehouse]:

* **No atomicity.** A writer that must add ten files has no way to make the ten
  appear together. Readers see a partial commit.
* **Listing is the query plan.** Planning cost scales with the number of files
  and, on object storage, each listing is a network round trip with eventual
  consistency behind it.
* **Schema is a directory convention.** Renaming a column, or changing how a
  table is partitioned, means rewriting paths, so in practice nobody does it.

Delta Lake [@armbrust2020delta] and Iceberg [@icebergspec] answer the same
question in the same shape — keep the file list in metadata, commit by swapping a
pointer — and differ in how that metadata is organised. A comparison of the three
main implementations, with a benchmark, is in @jain2023lakehouse.

## What Iceberg keeps instead

Iceberg's metadata is a tree, and every level of it is immutable:

| Level | Holds | Written as |
| --- | --- | --- |
| Table metadata | current schema, partition specs, snapshot log, properties | JSON |
| Snapshot | one manifest list — the state of the table at one instant | referenced from the metadata |
| Manifest list | the manifests in this snapshot, with partition ranges | Avro |
| Manifest | data files, with per-column bounds, null counts and row counts | Avro |
| Data file | the rows | Parquet, ORC or Avro |

A commit writes new metadata and then atomically swaps the pointer to it. Nothing
is mutated, which is why a reader mid-commit sees the old table rather than half
of the new one, and why yesterday's snapshot is still readable today: it is still
there, and still consistent.

`icebergr` exposes each level. The schema and partition spec come from the table
metadata:

```{r}
tbl <- icebergr_example_table(rows = 200)

icebergr_schema(tbl)
icebergr_partitions(tbl)
```

The snapshot log is the history:

```{r}
icebergr_snapshots(tbl)[, c("snapshot_id", "operation", "added_records")]
```

And the manifests are what makes a scan plan possible without opening any data
file — `icebergr_scan_plan()` reads bounds out of the manifest, not out of
Parquet:

```{r}
icebergr_scan_plan(icebergr_scan(tbl))[, c("data_file_path", "record_count")]
```

## Why the column statistics matter so much

Because per-column bounds live in the manifest, a filter can eliminate a file
before it is opened. That is the whole performance argument, and it is inherited
twice over: once from the manifest, and once from Parquet's own footer, whose
row-group statistics allow the same trick within a file.

The columnar layout underneath is Dremel's [@melnik2010dremel] — repetition and
definition levels, which is how a nested `struct` or `list` survives being stored
one column at a time. The gains from reading only the columns a query needs, and
only the row ranges that can match, are the ones measured for column stores
generally in @abadi2008column, and re-measured for Parquet and ORC specifically,
with their modern encodings, in @zeng2023columnar.

`vignette("pushdown")` shows what `icebergr` pushes down and how to confirm it
happened.

## Concurrency, without a lock server

Iceberg commits optimistically: a writer reads the current metadata, prepares a
new snapshot, and swaps the pointer only if the pointer has not moved. If it has,
the commit fails and is retried. All three lakehouse formats work this way
[@jain2023lakehouse].

What a reader gets from this is snapshot isolation in the sense of
@berenson1995critique: a scan runs entirely against one snapshot, so a
long-running read never observes a write that landed halfway through it. In
`icebergr` this is explicit rather than incidental, because a table handle *is* a
snapshot. Another session's commit is invisible until you ask for it:

```{r, eval = FALSE}
tbl <- icebergr_reload(tbl) # now points at the newest snapshot
```

That is a feature, not a staleness bug — two scans off the same handle are
guaranteed to agree.

## Where R sits in this

The design that makes an R client possible at all is the one described in
@pedreira2023composable: a data system decomposed into reusable components with
Arrow as the interchange format between them, rather than a monolith with one
front end. `icebergr` uses three of those components — `iceberg-rust`
[@icebergrust] for metadata and planning, Parquet readers underneath it, and the
Arrow C stream interface [@arrowcdata] to hand the result to R — and adds no
execution engine of its own.

The alternative route, reading Iceberg through DuckDB [@raasveldt2019duckdb], is a
good one for queries and remains available. What it cannot do is write, commit,
manage snapshots, or read the schema of a snapshot that is no longer current,
because those are table-format operations rather than query operations. Engines
whose whole design centres on this format — Photon, for instance
[@behm2022photon] — treat the metadata as a first-class input for exactly that
reason.

## What this package does not do

Iceberg's spec is larger than any single client implements, and this package is
narrower than `iceberg-rust`. `icebergr_spec_support()` reports the boundary for
your own build, and distinguishes the two cases — missing here, versus missing
upstream:

```{r}
support <- icebergr_spec_support()
features <- support$features
head(features[
  !is.na(features$supported) & !features$supported,
  c("feature", "reason")
], 8)
```

Reading a table another engine wrote with features this package does not
implement still works, as long as the scan does not need them. That is the
property the format was designed to have.

# References
