---
title: "Time Travel: Snapshots, Timestamps and Rollbacks"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
bibliography: vignette_reference.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time Travel: Snapshots, Timestamps and Rollbacks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(icebergr)
```

Every Iceberg commit leaves the previous state of the table intact and readable
[@icebergspec]. Nothing is overwritten, so a snapshot from before a bad load is
still there — and reading it needs no backup, no restore and no cooperation from
whoever wrote it.

```{r}
tbl <- icebergr_example_table(rows = 200)
```

## The snapshot log

```{r}
history <- icebergr_snapshots(tbl)
history[, c(
  "snapshot_id", "parent_snapshot_id", "operation", "added_records",
  "total_records"
)]
```

The example table was built with two appends, so there are two snapshots and the
second's parent is the first. `total_records` is cumulative; `added_records` is
what that commit contributed.

**Snapshot ids are `character`, not numeric.** Iceberg assigns them as random
64-bit integers and an R double carries only 53 bits of mantissa, so a numeric
round trip would silently select a *different, existing* snapshot rather than
fail. This is the one type decision in the package that exists purely to prevent
a wrong answer:

```{r}
class(history$snapshot_id)
```

## Reading a snapshot by id

```{r}
first <- history$snapshot_id[[1]]

nrow(icebergr_collect(icebergr_scan(tbl, snapshot_id = first)))
nrow(icebergr_collect(icebergr_scan(tbl)))
```

The first snapshot holds one append; the current table holds both. Pushdown works
against a historical snapshot exactly as it does against the current one, because
the manifests of that snapshot are still on disk with their statistics intact:

```{r}
nrow(icebergr_scan_plan(icebergr_scan(tbl, snapshot_id = first)))
nrow(icebergr_scan_plan(icebergr_scan(tbl)))
```

## Reading as of a time

`as_of` takes a time and resolves it to a snapshot:

```{r}
icebergr_collect(
  icebergr_scan(tbl, as_of = history$timestamp[[1]], select = "id", limit = 3)
)
```

Two things about that argument are worth being explicit about, because both are
easy to get wrong.

**It is a commit time, not a time in your data.** `as_of` asks "what did this
table look like then", not "which rows are from then". A `day` or `recorded_at`
column is data; filter on it with `filter`.

**It resolves against the snapshot log, not the snapshot list.** Iceberg keeps
both: the list of every snapshot that still exists, and the *log* of which
snapshot was current at which time. A rollback removes an entry from the log while
leaving the snapshot in the list. Resolving against the list would find an
abandoned snapshot with a matching timestamp and read a state the table was
rolled back *out of*; `icebergr` resolves against the log, so it does not.

The same applies to a snapshot that only ever existed on another branch — present
in the list, never in this branch's log, therefore never selected by `as_of`. It
is still reachable by id, which is the right split: an explicit id is a request
for a specific state, while a time is a request for the state this table was in.

## The schema travels too

Iceberg records a schema per snapshot, so a column that has since been renamed or
dropped is still nameable as of the snapshot that had it:

```{r}
identical(
  icebergr_schema(tbl, snapshot_id = first)$name,
  icebergr_schema(tbl)$name
)
```

Nothing was renamed here, so the two agree. Where they would not, `filter` and
`select` are resolved against the schema of the snapshot actually being read —
not the current one. That is what makes a historical read reproducible: the names
that worked then still work, and you do not have to know what happened to the
schema in between.

## Isolation, and why a handle looks stale

A table handle is bound to a snapshot. Another session's commit does not change
what your handle reads:

```{r}
current <- icebergr_snapshots(tbl)
tbl2 <- icebergr_reload(tbl)
identical(nrow(icebergr_snapshots(tbl2)), nrow(current))
```

`icebergr_reload()` is how you opt in to seeing new commits. Until you call it,
every scan off that handle reads one consistent state — snapshot isolation in the
sense of @berenson1995critique, obtained from immutability rather than from
locking. All three lakehouse formats get it the same way [@jain2023lakehouse].

The practical consequence: two scans off the same handle are guaranteed to agree,
which is what you want when a report runs several queries and has to add up.

## What is not here

Snapshot expiry — deleting old snapshots and the files only they reference — is a
maintenance operation this version does not expose, and neither is rollback
itself. Both need transaction actions `iceberg-rust` 0.10.0 does not provide:

```{r}
features <- icebergr_spec_support()$features
features[
  grepl("snapshot|expiry|rollback", features$feature, ignore.case = TRUE),
  c("feature", "supported", "reason")
]
```

Reading a table that another engine has rolled back or expired works normally.
Time travel is a read capability, and this package has all of it.

# References
