---
title: "Types: The R, Arrow and Iceberg Correspondence"
author: "Youzhi Yu<br><span style='font-size:85%;'>University of Chicago</span>"
bibliography: vignette_reference.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Types: The R, Arrow and Iceberg Correspondence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(icebergr)

# bit64 and vctrs are in Suggests; guard the chunks that need them so this
# vignette builds on any installation.
has_bit64 <- requireNamespace("bit64", quietly = TRUE)
has_vctrs <- requireNamespace("vctrs", quietly = TRUE)
```

Data crosses between R and Iceberg over the Arrow C stream interface
[@arrowcdata]: `iceberg-rust` writes Arrow arrays, `nanoarrow` [@nanoarrow]
reads the same memory, and nothing is serialised in between. Two type systems
still have to be reconciled at each end, and this vignette is about where that
reconciliation is exact and where it is not.

```{r}
warehouse <- file.path(tempdir(), "icebergr-types")
dir.create(warehouse, showWarnings = FALSE)
catalog <- icebergr_catalog("memory", warehouse = warehouse)
icebergr_create_namespace(catalog, "t")
```

## What survives unchanged

```{r}
plain <- data.frame(
  i = 1L,
  d = 1.5,
  s = "text",
  b = TRUE,
  day = as.Date("2026-07-01"),
  ts = as.POSIXct("2026-07-01 09:00:00", tz = "UTC")
)

tbl <- icebergr_create_table(catalog, "t.plain", plain)
icebergr_schema(tbl)[, c("name", "type")]
```

`integer`, `double`, `character`, `logical`, `Date` and `POSIXct` map onto `int`,
`double`, `string`, `boolean`, `date` and `timestamptz`, and come back as
themselves:

```{r}
tbl <- icebergr_append(tbl, plain)
back <- icebergr_collect(tbl)
vapply(back, function(x) class(x)[[1]], character(1))
identical(back$s, plain$s)
identical(back$day, plain$day)
```

A `POSIXct` is normalised to UTC — the instant is preserved, the printed time zone
is not. That is Iceberg's `timestamptz` semantics, not a lossy conversion.

## The five that need care

### `NaN` comes back as `NA`

```{r}
nan_tbl <- icebergr_create_table(catalog, "t.nan", data.frame(x = 1.5))
nan_tbl <- icebergr_append(nan_tbl, data.frame(x = c(1.5, NaN, NA)))
icebergr_collect(nan_tbl)$x
```

R's `is.na(NaN)` is `TRUE`, so `NaN` is written as a null and reads back as `NA`.
Nothing in the stack distinguishes them on the way out. If the difference matters,
encode it in another column.

### `factor` comes back as `character`

```{r}
fac_tbl <- icebergr_create_table(
  catalog, "t.fac",
  data.frame(g = factor("a", levels = c("a", "b")))
)
fac_tbl <- icebergr_append(fac_tbl, data.frame(g = factor(c("a", "b", "a"))))
str(icebergr_collect(fac_tbl)$g)
```

Iceberg has no dictionary type, so the levels have nowhere to live. Restore them
in R with `factor()` and your own level order — which is the honest outcome,
because a round trip that invented a level order would be worse.

### `long` needs `bit64`

Iceberg's `long` is 64-bit. An R double holds 53 bits of mantissa, so a value past
$2^{53}$ cannot survive as a `numeric`. With `bit64` installed it comes back
exact, at any depth — including inside a `struct`:

```{r, eval = has_bit64}
big <- bit64::as.integer64("9007199254740993") # 2^53 + 1
long_tbl <- icebergr_create_table(catalog, "t.long", data.frame(n = big))
long_tbl <- icebergr_append(long_tbl, data.frame(n = big))

got <- icebergr_collect(long_tbl)$n
class(got)
as.character(got)
```

```{r, eval = has_bit64}
identical(as.character(got), "9007199254740993")
```

Without `bit64`, Arrow's `int64` narrows to a `double` and that value reads back
as 9007199254740992. The package does not error on this — it is Arrow's
documented fallback — so if you handle 64-bit keys, put `bit64` in your own
`Imports`.

### `timestamp_ns` loses sub-microsecond precision

A nanosecond timestamp is readable, writable and filterable, but a `POSIXct` is a
double of *seconds*, so the bottom digits cannot be represented. `nanoarrow`
warns on every such read: the warning triggers on the nanosecond count exceeding
$2^{53}$, which any present-day instant does, so it appears even when nothing was
actually lost.

### Snapshot ids are `character`

Not a column type, but the same arithmetic. See `vignette("time-travel")`.

## Nested types

`struct` arrives as a data frame column, which is the natural R shape for it:

```{r}
nested <- data.frame(id = 1L)
nested$geo <- data.frame(lat = 41.79, lon = -87.6)

nest_tbl <- icebergr_create_table(catalog, "t.nested", nested)
icebergr_schema(nest_tbl)[, c("name", "type")]
```

```{r}
nest_tbl <- icebergr_append(nest_tbl, nested)
got <- icebergr_collect(nest_tbl)
class(got$geo)
got$geo$lat
```

A `list` column round-trips as a `vctrs::list_of`:

```{r, eval = has_vctrs}
lst <- data.frame(id = 1L)
lst$tags <- vctrs::list_of(c("a", "b"))

lst_tbl <- icebergr_create_table(catalog, "t.lst", lst)
lst_tbl <- icebergr_append(lst_tbl, lst)
icebergr_collect(lst_tbl)$tags
```

That nesting survives a columnar layout at all because of Dremel's repetition and
definition levels [@melnik2010dremel], which is how Parquet stores a repeated or
optional field one column at a time.

**Iceberg cannot push a filter or a projection *onto* a nested field.** Read the
parent column and subset in R:

```{r}
subset(icebergr_collect(nest_tbl), geo$lat > 41)
```

A `map` column can be created, but writing map *values* from R needs the `arrow`
package, because `nanoarrow` cannot build a map array on its own. If you do,
`nanoarrow::na_map()` needs its key type built non-nullable —
`na_map(na_string(nullable = FALSE), …)`.

## Reading the schema rather than guessing it

`icebergr_schema()` reports Iceberg's own types, with the field ids that make
schema evolution possible in the format:

```{r}
icebergr_schema(nest_tbl)
```

Field ids, not names, are what a manifest refers to — which is why Iceberg can
rename a column without rewriting data, and why the Hive convention could not
[@icebergspec; @thusoo2009hive].

## Why the bridge is an Arrow bridge

Handing R a pointer to Arrow memory that Rust allocated, with no copy and no
serialisation, is the concrete benefit of the composable-systems design
[@pedreira2023composable]: the interchange format is the contract, so neither side
needs to know the other's internals. The R half is `nanoarrow` [@nanoarrow]; the
binding itself is `extendr` [@extendr]; and the type correspondence above is the
whole of what this package has to get right at the boundary.

# References
